---
title: "Bellabeat Case Study in R"
author: "Silas Olumuyiwa Bamidele"
date: "2023-09-11"
output: html_document
---

## A. About Bellabeat

Bellabeat manufactures high-tech health-focused smart products.
Its Co-founder and artist Sršen has helped develop beautiful designed tech that empowers and inspires women all over the world. Bellabeat collects data on women activity, sleep, stress, and reproductive health.
Founded in 2013, Bellabeat has steadily and quickly grown to position itself as a wellness tech company for women around the world.

By 2016, Bellabeat had opened offices around the world and launched multiple products. Bellabeat products are available on Bellabeat.com and other online retail shops.

### Business Questions

1. What are some trends in smart device usage?
2. How could these trends apply to Bellabeat customers?
3. How could these trends help influence Bellabeat marketing strategy?

### Business task

To identify market opportunities for growth and provide high-level recommendations to Bellabeat to help guide the company's marketing strategy based on trends in smart device usage.

## B. Preparing and Cleaning Data

The FitBit Fitness Tracker Dataset has been recommended by Bellabeat's Co-founder and Chief Creative Officer, Urška Sršen.
### Content
This dataset generated by respondents to a distributed survey via Amazon Mechanical Turk between 03.12.2016-05.12.2016. Thirty eligible Fitbit users consented to the submission of personal tracker data, including minute-level output for physical activity, heart rate, and sleep monitoring.

#### Acknowledgements
The FitBit Fitness Tracker Dataset has been pulled from Kaggle
https://www.kaggle.com/datasets/arashnic/fitbit

#### Credits:  Furberg, Robert; Brinton, Julia; Keating, Michael ; Ortiz, Alexa

## B2. Exploratory Data Analysis
### B.2.1 Daily Activity Merged DF

I'm choosing to do this analysis in R because R affords me the sharing my notebook easily with colleagues and it's also easily reproduced and pulled from GitHub repositories.

First off, I'll be importing three (3) files from the dataset; 
* dailyActivity_merged
* hourlyCalories_merged
* sleepDay_merged

```{r }
library(readr)
library(dplyr)
library(tidyr)
library(stringr)

daily_activity <- read_csv("/home/silas/Desktop/Bellabeat/dailyActivity_merged.csv")
head(daily_activity) 


```
Fetching data summaries of the data and checking for missing values  
```{r}
summary(daily_activity)
dplyr::glimpse(daily_activity)
skimr::skim(daily_activity)

```
A quick inference from the summaries pulled here;
*For the dailyActivity_merged DF, there are 940 observations and 15 variables
* 1 charater column and 14 numeric columns, with the character column being dates which should be converted to a date type.
* No missing values in the dailyActivity_merged DF

### Convert character column to date type

```{r}
library(dplyr)
library(tidyr)
library(stringr)

daily_activity <- daily_activity %>%
  mutate(new_activity_date = as.Date(ActivityDate, format = "%m/%d/%Y")) %>% 
  mutate(day_of_week = weekdays(new_activity_date)) %>% 
  mutate(People_ID = as.character(Id))

#View(daily_activity) confirms the new_activity_date contains the dates in Date format
# Or the class(daily_activity$new_activity_date) at the console returns Date as the format
# NOTICE here, day of the week column is also created
```
  

### B.2.2 moving on to the hourlyCalories_merged dataframe;

```{r}
library(readr)
library(dplyr)
library(tidyr)
library(stringr)

hourly_calories <- read_csv("/home/silas/Desktop/Bellabeat/hourlyCalories_merged.csv")
head(hourly_calories) #fetching first 6 rows
anyNA(hourly_calories) #checking for missing values
summary(hourly_calories)

```
The summary shows there are no missing values on the hourly_calories DF.
Summary also shows the minimum calories burnt is 42 and the max is 948 calories.

### Now moving on to compute new columns for the hourly_calories Df;

```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
library(corrr)
library(plotly)

hourly_calories <- hourly_calories %>%
  mutate(activity_date = as.Date(ActivityHour, format = "%m/%d/%Y"))%>% 
  mutate(day_of_week = weekdays(activity_date)) %>% 
  mutate(People_ID = as.character(Id))

hourly_calories 

```

### B.2.3 Now to the sleepDay_merged DF
```{r}
library(readr)
library(dplyr)
library(tidyr)
library(stringr)

sleepday <- read_csv("/home/silas/Desktop/Bellabeat/sleepDay_merged.csv")
head(sleepday) 
anyNA(sleepday) # retuns false -- No missing values


```
## C. Visualizations and Deductions
### C1. Tracking calories burnt per participant

```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
library(corrr)
library(plotly)

hourly_calories <- hourly_calories %>%
  mutate(activity_date = as.Date(ActivityHour, format = "%m/%d/%Y"))%>% 
  mutate(day_of_week = weekdays(activity_date)) %>% 
  mutate(People_ID = as.character(Id))

hourly_calories %>% 
  ggplot(aes(x = People_ID, y = Calories, color = day_of_week)) +
  geom_col() + coord_flip() + theme_minimal() -> plot

ggplotly(plot)
```
### Mean in Calories
```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
library(corrr)
library(plotly)

hourly_calories %>% 
  group_by(day_of_week) %>% 
  summarise(mean_in_calories = mean(Calories, na.rm = TRUE)) %>% 
  fashion()  
```
This suggests there's good correlation in the calculated means in calories burnt -- an indication that the participants used the Hourly Calories app everyday for the 3 months period of activity.
It also shows participants rested a little bit more on Sundays.

### C2. Total minutes slept 
```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
library(corrr)
library(plotly)

# Converting Id column to People_ID character format and creating new computed columns, day_of_week and activity_date
sleepday <- sleepday %>%
  mutate(activity_date = as.Date(SleepDay, format = "%m/%d/%Y"))%>% 
  mutate(day_of_week = weekdays(activity_date)) %>% 
  mutate(People_ID = as.character(Id))

sleepday %>% 
  ggplot(aes(x = People_ID, y = TotalMinutesAsleep, color = day_of_week)) +
  geom_col() + coord_flip() + theme_minimal() -> plot

ggplotly(plot)

```

### Computing the average sleep in minutes
```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
library(corrr)
library(plotly)

sleepday %>% 
  group_by(day_of_week) %>% 
  summarise(mean_TotalMinutesAsleep = mean(TotalMinutesAsleep, na.rm = TRUE)) %>% 
  fashion() 

```
This validates participants slept more on Sundays during the 3-month period as average in minutes of sleep is highest on Sunday.

### C3. Daily Activity Plots
#### C.3.1 Very Active Minutes VS Calories

```{r}
library(ggplot2)
library(dplyr)
library(tidyr)
library(corrr)
library(plotly)

daily_activity %>% 
  ggplot(aes(x = VeryActiveMinutes, y = Calories, color = day_of_week)) +
  geom_jitter() + geom_abline() + theme_minimal() +
  facet_wrap(~ People_ID) -> plot

ggplotly(plot)
```
This plot shows a linear relationship between VeryActiveMinutes and Calories 

#### C.3.2 Sedentary Minutes VS Calories

```{r echo=FALSE}

library(ggplot2)
library(dplyr)
library(tidyr)
library(corrr)
library(plotly)

daily_activity %>% 
  ggplot(aes(x = SedentaryMinutes, y = Calories, color = day_of_week)) +
  geom_jitter() + theme_minimal() +
  facet_wrap(~ People_ID) -> plot

ggplotly(plot)
```
This sure shows a decline in calories burnt as sedentary minutes increases per participant.

#### C.3.3 Total Steps VS Calories

```{r warning=FALSE}
library(ggplot2)
library(dplyr)
library(tidyr)
library(plotly)

daily_activity %>% 
  ggplot(aes(x = TotalSteps, y = Calories, color = People_ID)) +
  geom_point() + geom_smooth() + theme_minimal() +
  facet_wrap(~ day_of_week) -> plot

ggplotly(plot)

```
Increase in calories burnt as total steps increase

## 4. Conclusions

1. Trends show that participants use the apps less on Sundays.
2. As sedentary minutes increases, calories burnt decreases.
3. The more active the participants, the more calories they burn.
4. Survey participants did not all participate at the same level, with some very active and some not so active using the apps.
5. Include feedback forms in-app to see what apps customers would use more and see what the reasons are.

## 5. Recommendations
Having been tested by 30 participants in a 3-month survey and with the results seen, the FitBit Fitness Tracker is recommended as a benchmark to building and/or extending Bellabeat's own fitness apps.
Some cues to note are;
* Getting more customers to use the apps everyday. Building friendly notifiers and exercise tracking plans would go a long way here.
* Including reward systems in apps to help customers use all the apps in Bellabeat is also recommended.

Thank You.
Silas O. Bamidele


